---
title: "Stop and Frisk in New York City from 2017 - 2019"
# listing: texts
description: |
  An analysis of stop and frisk plicy in NYC using R. This article is based off a project initially completed in python. 
author:
  - name: Roupen Khanjian
    url: {}
date: 03-14-2021
categories:
  - Text Analysis
  - R
  - Python
output:
  distill::distill_article:
    # self_contained: false
    # theme: darkly
    # css: stylesNBA.css
    code_folding: Code
    toc: true
    # toc_float: true # didnt work
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

<style>
pre.r {
    background-color: #e7eaea;
}
</style>

```{r}
# Packages 
library(tidyverse) # Easily Install and Load the 'Tidyverse', CRAN v1.3.0
library(patchwork) # The Composer of Plots, CRAN v1.1.1
library(here) # A Simpler Way to Find Your Files, CRAN v1.0.1
library(janitor) # Simple Tools for Examining and Cleaning Dirty Data, CRAN v2.1.0
library(tidytext) # Text Mining using 'dplyr', 'ggplot2', and Other Tidy Tools, CRAN v0.3.0
library(textdata) # Download and Load Various Text Datasets, CRAN v0.4.1
library(vader) # Valence Aware Dictionary and sEntiment Reasoner (VADER), CRAN v0.2.1
library(readxl) # Read Excel Files, CRAN v1.3.1
library(lubridate) # Make Dealing with Dates a Little Easier, CRAN v1.7.10
library(gt) # Easily Create Presentation-Ready Display Tables, CRAN v0.2.2
library(kableExtra) # Construct Complex Table with 'kable' and Pipe Syntax, CRAN v1.3.4
library(knitr) # A General-Purpose Package for Dynamic Report Generation in R, CRAN v1.31
library(plotly) # Create Interactive Web Graphics via 'plotly.js', CRAN v4.9.3

```

## Introduction

This article was originally completed in python here: [link](https://khanjian.github.io/roupen-website/stop_and_frisk_python.html)

During the past 3 weeks, protests have taken place all over the country and the world in response to the murder of George Floyd by the Minneapolis Police. Since the advent of video recordings via cell phones, there have been many instances of police brutality captured against Black people. The relationship between law enforcement and the Black community has always been tenuous and now it has been brought to the attention of the rest of the world. In the landmark Supreme court case of Terry v. Ohio (1968) it was ruled that police could stop, question, and frisk a person if they have reasonable suspicion that the person had committed a crime (Brandes, S.A. et al., 2019).

In the 2000’s to the early 2010’s the New York City (NYC) stop and frisk policy garnered national attention due to the high number of stops and profiling of Black people. At the height of the policy, in 2011 there were 658,724 stops recorded with over 50% of the stops targeting Black people (NYCLU 2019). Since then, the number of stops per year has substantially decreased to 13,459 stop in 2019. Opponents of this policy argue this is still too many stops, especially since in 2019 about 66% of the people stopped were innocent. It has been also shown that the stopping of white people more likely led to an arrest in comparison to Black and Hispanic people, implying the police may be targeting minorities and being more mindful of stopping white people (Gelman, A., et. al., 2007). There has been substantial research conducted showing the psychological distress of a stop and frisk policy on communities of color in NYC (Sewell, A. et al., 2016).

In this project I analyze the Stop, Question and Frisk Data from the New York Police Department (NYPD) from the most current three years: 2017, 2018, and 2019 (NYC Stop and Frisk Data). I chose these years for the following reasons: The years 2018-2019 was not included in the most recent NYCLU report. In 2017 the NYPD moved to an electronic form, as opposed to manually writing down a response for each question in the handwritten forms used prior to 2017. Lastly 2017 was the first year of the Trump presidency and I was curious to investigate if his rhetoric on race may have affected law enforcements behavior toward minorities.

Data was downloaded from the NYPD website [link](https://www1.nyc.gov/site/nypd/stats/reports-analysis/stopfrisk.page). In each dataset each row is a stop of a specific person, and each column is a variable. There are a total of 83 different variables in each dataset.

```{r}
# Read in data
# file_names <- list.files(here("_texts", 
#                               "stop_frisk_R", 
#                               "data")) # file names for each episode
# 
# frisk <- str_glue(here("_texts","stop_frisk_R", 
#                               "data/"),"{file_names}") %>% 
#   map_dfr(read_excel) # read in all the episodes into one data frame!

saf_17 <- read_excel(here("_texts", "stop_frisk_R", "data",
                          "sqf_2017.xlsx"))

saf_18 <- read_excel(here("_texts", "stop_frisk_R", "data",
                          "sqf_2018.xlsx"))

saf_19 <- read_excel(here("_texts", "stop_frisk_R", "data",
                          "sqf_2019.xlsx"))

# fix date column and select columns to be used for analysis
saf_17 <- saf_17 %>%
  mutate(STOP_FRISK_DATE = ymd(STOP_FRISK_DATE)) %>%
  select(
    c(
      "STOP_FRISK_DATE",
      "STOP_FRISK_TIME",
      "YEAR2",
      "MONTH2",
      "DAY2",
      "ISSUING_OFFICER_RANK",
      "SUSPECTED_CRIME_DESCRIPTION",
      "SUSPECT_ARRESTED_FLAG",
      "SUSPECT_ARREST_OFFENSE",
      "OFFICER_IN_UNIFORM_FLAG",
      "FRISKED_FLAG",
      "SEARCHED_FLAG",
      "OTHER_CONTRABAND_FLAG",
      "FIREARM_FLAG",
      "KNIFE_CUTTER_FLAG",
      "OTHER_WEAPON_FLAG",
      "WEAPON_FOUND_FLAG",
      "PHYSICAL_FORCE_CEW_FLAG",
      "PHYSICAL_FORCE_DRAW_POINT_FIREARM_FLAG",
      "PHYSICAL_FORCE_HANDCUFF_SUSPECT_FLAG",
      "PHYSICAL_FORCE_OC_SPRAY_USED_FLAG",
      "PHYSICAL_FORCE_OTHER_FLAG",
      "PHYSICAL_FORCE_RESTRAINT_USED_FLAG",
      "PHYSICAL_FORCE_VERBAL_INSTRUCTION_FLAG",
      "PHYSICAL_FORCE_WEAPON_IMPACT_FLAG",
      "SUSPECTS_ACTIONS_CASING_FLAG",
      "SUSPECTS_ACTIONS_PROXIMITY_TO_SCENE_FLAG",
      "DEMEANOR_OF_PERSON_STOPPED",
      "SUSPECT_REPORTED_AGE",
      "SUSPECT_SEX",
      "SUSPECT_RACE_DESCRIPTION",
      "SUSPECT_BODY_BUILD_TYPE",
      "SUSPECT_OTHER_DESCRIPTION",
      "STOP_LOCATION_PRECINCT",
      "STOP_LOCATION_FULL_ADDRESS",
      "STOP_LOCATION_STREET_NAME",
      "STOP_LOCATION_PATROL_BORO_NAME",
      "STOP_LOCATION_BORO_NAME",
      "SUSPECTS_ACTIONS_DRUG_TRANSACTIONS_FLAG"
    )
  )

saf_18 <- saf_18 %>%
  mutate(STOP_FRISK_DATE = ymd(STOP_FRISK_DATE)) %>%
  select(
    c(
      "STOP_FRISK_DATE",
      "STOP_FRISK_TIME",
      "YEAR2",
      "MONTH2",
      "DAY2",
      "ISSUING_OFFICER_RANK",
      "SUSPECTED_CRIME_DESCRIPTION",
      "SUSPECT_ARRESTED_FLAG",
      "SUSPECT_ARREST_OFFENSE",
      "OFFICER_IN_UNIFORM_FLAG",
      "FRISKED_FLAG",
      "SEARCHED_FLAG",
      "OTHER_CONTRABAND_FLAG",
      "FIREARM_FLAG",
      "KNIFE_CUTTER_FLAG",
      "OTHER_WEAPON_FLAG",
      "WEAPON_FOUND_FLAG",
      "PHYSICAL_FORCE_CEW_FLAG",
      "PHYSICAL_FORCE_DRAW_POINT_FIREARM_FLAG",
      "PHYSICAL_FORCE_HANDCUFF_SUSPECT_FLAG",
      "PHYSICAL_FORCE_OC_SPRAY_USED_FLAG",
      "PHYSICAL_FORCE_OTHER_FLAG",
      "PHYSICAL_FORCE_RESTRAINT_USED_FLAG",
      "PHYSICAL_FORCE_VERBAL_INSTRUCTION_FLAG",
      "PHYSICAL_FORCE_WEAPON_IMPACT_FLAG",
      "SUSPECTS_ACTIONS_CASING_FLAG",
      "SUSPECTS_ACTIONS_PROXIMITY_TO_SCENE_FLAG",
      "DEMEANOR_OF_PERSON_STOPPED",
      "SUSPECT_REPORTED_AGE",
      "SUSPECT_SEX",
      "SUSPECT_RACE_DESCRIPTION",
      "SUSPECT_BODY_BUILD_TYPE",
      "SUSPECT_OTHER_DESCRIPTION",
      "STOP_LOCATION_PRECINCT",
      "STOP_LOCATION_FULL_ADDRESS",
      "STOP_LOCATION_STREET_NAME",
      "STOP_LOCATION_PATROL_BORO_NAME",
      "STOP_LOCATION_BORO_NAME",
      "SUSPECTS_ACTIONS_DRUG_TRANSACTIONS_FLAG"
    )
  )

saf_19 <- saf_19 %>%
  mutate(STOP_FRISK_DATE = ymd(STOP_FRISK_DATE)) %>%
  select( # select for the years 
    c(
      "STOP_FRISK_DATE",
      "STOP_FRISK_TIME",
      "YEAR2",
      "MONTH2",
      "DAY2",
      "ISSUING_OFFICER_RANK",
      "SUSPECTED_CRIME_DESCRIPTION",
      "SUSPECT_ARRESTED_FLAG",
      "SUSPECT_ARREST_OFFENSE",
      "OFFICER_IN_UNIFORM_FLAG",
      "FRISKED_FLAG",
      "SEARCHED_FLAG",
      "OTHER_CONTRABAND_FLAG",
      "FIREARM_FLAG",
      "KNIFE_CUTTER_FLAG",
      "OTHER_WEAPON_FLAG",
      "WEAPON_FOUND_FLAG",
      "PHYSICAL_FORCE_CEW_FLAG",
      "PHYSICAL_FORCE_DRAW_POINT_FIREARM_FLAG",
      "PHYSICAL_FORCE_HANDCUFF_SUSPECT_FLAG",
      "PHYSICAL_FORCE_OC_SPRAY_USED_FLAG",
      "PHYSICAL_FORCE_OTHER_FLAG",
      "PHYSICAL_FORCE_RESTRAINT_USED_FLAG",
      "PHYSICAL_FORCE_VERBAL_INSTRUCTION_FLAG",
      "PHYSICAL_FORCE_WEAPON_IMPACT_FLAG",
      "SUSPECTS_ACTIONS_CASING_FLAG",
      "SUSPECTS_ACTIONS_PROXIMITY_TO_SCENE_FLAG",
      "DEMEANOR_OF_PERSON_STOPPED",
      "SUSPECT_REPORTED_AGE",
      "SUSPECT_SEX",
      "SUSPECT_RACE_DESCRIPTION",
      "SUSPECT_BODY_BUILD_TYPE",
      "SUSPECT_OTHER_DESCRIPTION",
      "STOP_LOCATION_PRECINCT",
      "STOP_LOCATION_FULL_ADDRESS",
      "STOP_LOCATION_STREET_NAME",
      "STOP_LOCATION_PATROL_BORO_NAME",
      "STOP_LOCATION_BORO_NAME",
      "SUSPECTS_ACTIONS_DRUG_TRANSACTIONS_FLAG"
    )
  )

# row bind all 3 years into one data frame
saf <- rbind(saf_17, saf_18, saf_19) 

# clean column names
saf <- saf %>% 
  clean_names()

# view first 5 rows
head(saf, n = 10) %>% 
  kbl(caption = "<b style = 'color:black;'>
       First ten rows of stop and frisk dataset.") %>%
  kable_material_dark(bootstrap_options = c("striped", "hover")) %>%
  row_spec(0, color = "white", background = "#222222") %>%
  scroll_box(width = "100%", height = "300px", 
             fixed_thead = list(enabled = T, background = "#222222"))

nrow(saf)
```

* In total there were 36,096 stops in the years 2017-2019.

```{r}

saf %>% 
  count(suspect_race_description) %>% 
  summarise(race = suspect_race_description,
            n = n,
            percent = n / sum(n)) %>% 
  gt() %>% 
  tab_header(
    title = "Which Race was Stop and Frisked the Most?",
    subtitle = "Data from 2017-2019"
  ) %>% 
  fmt_percent( 
    columns = vars(percent),
    decimals = 2
    ) %>% 
  data_color(
    columns = vars(n, percent),
    colors = scales::col_numeric(
      palette = c(
        "lightskyblue", "dodgerblue", "royalblue4") ,
      domain = NULL
        )
      ) %>% 
  tab_style(
    style = list(
      cell_text(style = "oblique"),
      cell_borders(
        side = c("right"), 
        color = "black",
        weight = px(2)
        )
    ),
    locations = cells_body(
      columns = 1
    )
    )

```

In the above table we see that the people stopped were mostly described as have one of the following four races: Black, White Hispanic, White, or Black Hispanic.

## Data Wrangling and Cleaning

```{r}

saf_sub <- saf %>% 
  # filter for race in one of the four most saf 
  filter(suspect_race_description %in% c("BLACK", 
                                         "WHITE",
                                         "WHITE HISPANIC",
                                         "BLACK HISPANIC")) %>% 
  # Fill in NA values for nonsense descriptions
  # also change misspellings
  mutate(suspect_other_description = case_when(
    suspect_other_description == "(null)" ~ NA_character_,
    suspect_other_description == "NONE" ~ NA_character_,
    suspect_other_description == "UNK" ~ NA_character_,
    suspect_other_description == "UNKNOWN" ~ NA_character_,
    suspect_other_description == "UNKNOW" ~ NA_character_,
    suspect_other_description == "NO" ~ NA_character_,
    suspect_other_description == "UKNOWN" ~ NA_character_,
    suspect_other_description == "UNKOWN" ~ NA_character_,
    suspect_other_description == "NA" ~ NA_character_,
    suspect_other_description == "N/A" ~ NA_character_,
    suspect_other_description == "0" ~ NA_character_,
    suspect_other_description == 0 ~ NA_character_,
    suspect_other_description == "TATOOS" ~ "TATTOOS",
    suspect_other_description == "TATTOS" ~ "TATTOOS",
    suspect_other_description == "TATOO" ~ "TATTOOS",
    TRUE ~ as.character(suspect_other_description)
  )) %>% 
  # Change unknown sex values to NA
  mutate(suspect_sex = case_when(
    suspect_sex == "(null)" ~ NA_character_,
    suspect_sex == "19" ~ NA_character_,
    suspect_sex == "23" ~ NA_character_,
    suspect_sex == "24" ~ NA_character_,
    suspect_sex == "39" ~ NA_character_,
    TRUE ~ as.character(suspect_sex)
  )) %>% 
  # Change (null) value to No (N) (need to find a better way to do this step)
  mutate(physical_force_cew_flag = case_when(
    physical_force_cew_flag == "(null)" ~ "N",
    TRUE ~ as.character(physical_force_cew_flag)
  )) %>% 
  mutate(physical_force_draw_point_firearm_flag = case_when(
    physical_force_draw_point_firearm_flag == "(null)" ~ "N",
    TRUE ~ as.character(physical_force_draw_point_firearm_flag)
  )) %>% 
  mutate(physical_force_handcuff_suspect_flag = case_when(
    physical_force_handcuff_suspect_flag == "(null)" ~ "N",
    TRUE ~ as.character(physical_force_handcuff_suspect_flag)
  )) %>% 
  mutate(physical_force_oc_spray_used_flag = case_when(
    physical_force_oc_spray_used_flag == "(null)" ~ "N",
    TRUE ~ as.character(physical_force_oc_spray_used_flag)
  )) %>% 
  mutate(physical_force_other_flag = case_when(
    physical_force_other_flag == "(null)" ~ "N",
    TRUE ~ as.character(physical_force_other_flag)
  )) %>% 
  mutate(physical_force_restraint_used_flag = case_when(
    physical_force_restraint_used_flag == "(null)" ~ "N",
    TRUE ~ as.character(physical_force_restraint_used_flag)
  )) %>% 
  mutate(physical_force_verbal_instruction_flag = case_when(
    physical_force_verbal_instruction_flag == "(null)" ~ "N",
    TRUE ~ as.character(physical_force_verbal_instruction_flag)
  )) %>% 
  mutate(physical_force_weapon_impact_flag = case_when(
    physical_force_weapon_impact_flag == "(null)" ~ "N",
    TRUE ~ as.character(physical_force_weapon_impact_flag)
  )) %>% 
  mutate(suspects_actions_drug_transactions_flag = case_when(
    suspects_actions_drug_transactions_flag == "(null)" ~ "N",
    TRUE ~ as.character(suspects_actions_drug_transactions_flag)
  )) %>% 
  mutate(suspects_actions_proximity_to_scene_flag = case_when(
    suspects_actions_proximity_to_scene_flag == "(null)" ~ "N",
    TRUE ~ as.character(suspects_actions_proximity_to_scene_flag)
  )) %>% 
  mutate(other_contraband_flag = case_when(
    other_contraband_flag == "(null)" ~ "N",
    TRUE ~ as.character(other_contraband_flag)
  )) %>% 
  mutate(firearm_flag = case_when(
    firearm_flag == "(null)" ~ "N",
    TRUE ~ as.character(firearm_flag)
  )) %>% 
  mutate(knife_cutter_flag = case_when(
    knife_cutter_flag == "(null)" ~ "N",
    TRUE ~ as.character(knife_cutter_flag)
  )) %>% 
  mutate(weapon_found_flag = case_when(
    weapon_found_flag == "(null)" ~ "N",
    weapon_found_flag == "(" ~ "N",
    TRUE ~ as.character(weapon_found_flag)
  )) %>% 
  mutate(suspect_arrest_offense = case_when(
    suspect_arrest_offense == "(null)" ~ "NO ARREST",
    TRUE ~ as.character(suspect_arrest_offense)
  )) %>% 
  mutate(stop_location_boro_name = case_when(
    stop_location_boro_name == "(null)" ~ NA_character_,
    stop_location_boro_name == "STATEN IS" ~ "STATEN ISLAND",
    stop_location_boro_name == "PBBX" ~ "BRONX",
    stop_location_boro_name == "PBBN" ~ "BROOKLYN",
    stop_location_boro_name == "PBMN" ~ "MANHATTAN",
    stop_location_boro_name == "0208760" ~ NA_character_,
    stop_location_boro_name == "0190241" ~ NA_character_,
    stop_location_boro_name == "0986759" ~ NA_character_,
    stop_location_boro_name == "PBMS" ~ "MANHATTAN",
    stop_location_boro_name == "0210334" ~ NA_character_,
    stop_location_boro_name == "PBSI" ~ "STATEN ISLAND",
    stop_location_boro_name == "0237177" ~ NA_character_,
    stop_location_boro_name == "PBBS" ~ "BROOKLYN",
    stop_location_boro_name == "0155070" ~ NA_character_,
    stop_location_boro_name == "0208169" ~ NA_character_,
    TRUE ~ as.character(stop_location_boro_name)
  )) %>% 
  mutate(stop_location_precinct = case_when(
    stop_location_precinct == "(null)" ~ NA_integer_,
    stop_location_precinct == 208760 ~ NA_integer_,
    TRUE ~ as.integer(stop_location_precinct)
  )) %>% 
  mutate(demeanor_of_person_stopped = case_when(
    demeanor_of_person_stopped == "IRRATE" ~ "IRATE",
    demeanor_of_person_stopped == "1" ~ NA_character_,
    demeanor_of_person_stopped == "NEVEVOUS" ~ "NERVOUS",
    demeanor_of_person_stopped == 1 ~ NA_character_,
    demeanor_of_person_stopped == "N/A" ~ NA_character_,
    TRUE ~ as.character(demeanor_of_person_stopped)
  )) %>% 
  mutate(suspect_reported_age = case_when(
    suspect_reported_age == "(null)" ~ NA_character_,
    suspect_reported_age == "0" ~ NA_character_,
    suspect_reported_age == "1" ~ NA_character_,
    TRUE ~ as.character(suspect_reported_age)
  )) %>% 
  mutate(suspect_reported_age = as.integer(suspect_reported_age))

```

Due to the nature of the dataset, there had to be some data cleaning conducted in order to organize some of the columns:

* For most of the columns that are flags, converted (null) or other values to `N` (No) values.

* Changed non male or female values to `NA` in the `suspect_sex` column.

* In the ``r str_to_lower("SUSPECT_OTHER_DESCRIPTION")``, classified null of other values that were obviously null values to `NA`.

* In the ``r str_to_lower("DEMEANOR_OF_PERSON_STOPPED")`` corrected some misspellings and classified some observations to `NA` if no demeanor was noted.

* Organized the ``r str_to_lower("STOP_LOCATION_BORO_NAME")`` so that it represents one of the 5 boroughs of NYC or `NA` if it was not clear how to classify the stop.

* Changed the (null) values for the ``r str_to_lower("SUSPECT_ARREST_OFFENSE")`` to "No Arrest".

* Finally, converted missing values in the ``r str_to_lower("STOP_LOCATION_PRECINCT")`` column to `NA` since the precincts are numbered from 1 to 123.

## Exploratory Data Analysis (EDA)

The cleaned and updated dataset includes 34,784 observations ommiting the stops conducting on people that were not described as Black, White, Black Hispanic, or White Hispanic.

```{r, layout="l-body-outset"}
saf_sub %>% 
  count(suspect_race_description) %>% 
  ggplot(aes(x = n, y = fct_reorder(suspect_race_description, n))) +
  geom_col(fill = "#7dc8c4") +
  labs(x = "Number of Stops",
       y = "Race",
       title = "Number of Stops by Race") +
  theme_minimal() +
  theme(panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank(),
        axis.text = element_text(size = 11,
                                   face = "bold"),
        axis.title.x = element_text(size = 14,
                                   face = "bold"),
        axis.title.y = element_blank(),
        panel.grid.major.x = element_line(color = "grey27"),
        panel.grid.minor.x = element_line(color = "grey27"),
        plot.background = element_rect(colour = "#e7eaea",
                                       fill = "#e7eaea"))
```

* In the above barplot, we see that Black people are stopped substantially more than any other group. They represent over 50% of the total stops, even though they make up 25% of the population in NYC (NYC census data 2019).

```{r, layout="l-body-outset"}
saf_sub %>% 
  count(suspect_sex) %>% 
  drop_na() %>% 
  ggplot(aes(x = n, y = fct_reorder(suspect_sex, n))) +
  geom_col(fill = "#7dc8c4") +
  labs(x = "Number of Stops",
       y = "Sex",
       title = "Number of Stops by Sex") +
  theme_minimal() +
  theme(panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank(),
        axis.text = element_text(size = 11,
                                   face = "bold"),
        axis.title.x = element_text(size = 14,
                                   face = "bold"),
        axis.title.y = element_blank(),
        panel.grid.major.x = element_line(color = "grey27"),
        panel.grid.minor.x = element_line(color = "grey27"),
        plot.background = element_rect(colour = "#e7eaea",
                                       fill = "#e7eaea"))
```

* The majority of stops are of males.

```{r, layout="l-body-outset"}

d1 <- saf_sub %>% 
  count(day2) %>% 
  ggplot(aes(x = n, y = fct_reorder(day2, n))) +
  geom_col(fill = "#7dc8c4") +
  labs(x = "Number of Stops",
       y = "Weekday",
       title = "Number of Stops by  Weekday") +
  theme_minimal() +
  theme(panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank(),
        axis.text = element_text(size = 11,
                                   face = "bold"),
        axis.title.x = element_text(size = 14,
                                   face = "bold"),
        axis.title.y = element_blank(),
        panel.grid.major.x = element_line(color = "grey27"),
        panel.grid.minor.x = element_line(color = "grey27"),
        plot.background = element_rect(colour = "#e7eaea",
                                       fill = "#e7eaea"))

m1 <- saf_sub %>% 
  count(month2) %>% 
  ggplot(aes(x = n, y = fct_reorder(month2, n))) +
  geom_col(fill = "#7dc8c4") +
  labs(x = "Number of Stops",
       y = "Month",
       title = "Number of Stops by Month") +
  theme_minimal() +
  theme(panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank(),
        axis.text = element_text(size = 11,
                                   face = "bold"),
        axis.title.x = element_text(size = 14,
                                   face = "bold"),
        axis.title.y = element_blank(),
        panel.grid.major.x = element_line(color = "grey27"),
        panel.grid.minor.x = element_line(color = "grey27"),
        plot.background = element_rect(colour = "#e7eaea",
                                       fill = "#e7eaea"))

y1 <- saf_sub %>% 
  count(year2) %>% 
  mutate(year = fct_reorder(factor(year2), n)) %>% 
  ggplot(aes(x = n, y = year)) +
  geom_col(fill = "#7dc8c4") +
  labs(x = "Number of Stops",
       y = "Month",
       title = "Number of Stops by Month") +
  theme_minimal() +
  theme(panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank(),
        axis.text = element_text(size = 11,
                                   face = "bold"),
        axis.title.x = element_text(size = 14,
                                   face = "bold"),
        axis.title.y = element_blank(),
        panel.grid.major.x = element_line(color = "grey27"),
        panel.grid.minor.x = element_line(color = "grey27"),
        plot.background = element_rect(colour = "#e7eaea",
                                       fill = "#e7eaea"))
(d1 + m1) / y1

```

* Over 5500 of the stops occurred on Saturdays, while less than 4000 stops occurred on Mondays.

* The least number of stops were made in December, while the Spring months of March, April and May had the highest number of stops.

```{r, layout="l-body-outset"}
saf_sub %>% 
  drop_na(suspect_reported_age) %>% 
  filter(suspect_reported_age >= 10 & 
           suspect_reported_age <= 80) %>% 
  ggplot(aes(x = suspect_reported_age)) +
  geom_histogram(fill = "#7dc8c4",color = "white", 
                 binwidth = 5) +
  scale_x_continuous(breaks = seq(10, 80, 10)) +
  labs(x = "Age",
       y = "Number of Stops",
       title = "Number of Stops by Month") +
  theme_minimal() +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        axis.text = element_text(size = 11,
                                   face = "bold"),
        axis.title.x = element_text(size = 14,
                                   face = "bold"),
        axis.title.y = element_blank(),
        panel.grid.major.y = element_line(color = "grey27"),
        panel.grid.minor.y = element_line(color = "grey27"),
        plot.background = element_rect(colour = "#e7eaea",
                                       fill = "#e7eaea"))
```

* Above is a histogram of stop organized by age. We see a right skewed distribution, where the majority of people stopped were between the ages 15-30. I omitted people that were aged 80 or more in this histogram since there were so few observations.


```{r, layout="l-body-outset"}

saf_sub %>% 
  count(stop_location_boro_name) %>%
  drop_na() %>% 
  ggplot(aes(x = n, y = fct_reorder(stop_location_boro_name, n))) +
  geom_col(fill = "#7dc8c4") +
  labs(x = "Number of Stops",
       y = "Borough",
       title = "Number of Stops by Borough") +
  theme_minimal() +
  theme(panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank(),
        axis.text = element_text(size = 11,
                                   face = "bold"),
        axis.title.x = element_text(size = 14,
                                   face = "bold"),
        axis.title.y = element_blank(),
        panel.grid.major.x = element_line(color = "grey27"),
        panel.grid.minor.x = element_line(color = "grey27"),
        plot.background = element_rect(colour = "#e7eaea",
                                       fill = "#e7eaea"))
  

```

* We see that the most number of stops is made in Brooklyn. This is expected since Brooklyn has the highest population of all the boroughs (NYC census data). It is interesting that the second most number of stops is made in Manhattan, even though Queens is the second most populous borough and has about 600,000 more residents. Many NYC residents commute to and work in Manhattan, so this could explain this discrepancy between number of stops and population of each borough.

```{r,  layout="l-body-outset"}

plotly1 <- saf_sub %>% 
  drop_na(suspect_reported_age, suspect_race_description) %>% 
  filter(suspect_reported_age >= 10 & 
           suspect_reported_age <= 80) %>% 
  count(suspect_race_description, suspect_reported_age) %>% 
  ggplot(aes(x = suspect_race_description, y = suspect_reported_age)) +
  geom_tile(aes(fill = n)) +
  scale_fill_viridis_c(option = "cividis") +
  scale_y_continuous(breaks = seq(10, 80, 5)) +
  labs(x = "Race") +
  theme_minimal() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text = element_text(size = 11,
                                   face = "bold"),
        axis.title.x = element_text(size = 14,
                                   face = "bold"),
        axis.title.y = element_blank(),
        plot.background = element_rect(colour = "#e7eaea",
                                       fill = "#e7eaea"))

ggplotly(plotly1)

```

* In the above interactive visualization we see that the highest subgroup of people stopped were Black people aged 16-17 (1107 stops). The highest subgroup of White people that were stopped were of the age 30-31 (210 stops).

```{r}

saf_sub %>% 
  count(suspect_race_description, suspected_crime_description) %>% 
  pivot_wider(names_from = suspect_race_description,
              values_from = n) %>% 
  gt() %>% 
  tab_header(
    title = "Race and Alleged Crime Description",
    subtitle = "Data from 2017-2019"
  )  %>% 
  data_color(
    columns = vars(BLACK,`BLACK HISPANIC`,
                   WHITE,`WHITE HISPANIC`),
    colors = scales::col_numeric(
      palette = c(
        "lightskyblue", "dodgerblue", "royalblue4") ,
      domain = NULL
        )
      ) %>% 
  tab_style(
    style = list(
      cell_text(style = "oblique"),
      cell_borders(
        side = c("right"), 
        color = "black",
        weight = px(2)
        )
    ),
    locations = cells_body(
      columns = 1
    )
    )

```

* In the above table we see the breakdown of the alleged crime a person had commited for each stopped by each race. We see for all four races criminal possesion of a weapon (CPW) as the numebr one reason for stopping someone.


```{r}

saf_sub %>% 
  count(suspect_race_description, suspected_crime_description) %>% 
  pivot_wider(names_from = suspect_race_description,
              values_from = n) %>% 
  gt() %>% 
  tab_header(
    title = "Race and Alleged Crime Description",
    subtitle = "Data from 2017-2019"
  )  %>% 
  data_color(
    columns = vars(BLACK,`BLACK HISPANIC`,
                   WHITE,`WHITE HISPANIC`),
    colors = scales::col_numeric(
      palette = c(
        "lightskyblue", "dodgerblue", "royalblue4") ,
      domain = NULL
        )
      ) %>% 
  tab_style(
    style = list(
      cell_text(style = "oblique"),
      cell_borders(
        side = c("right"), 
        color = "black",
        weight = px(2)
        )
    ),
    locations = cells_body(
      columns = 1
    )
    )

```


